Building and Checking a Suffix Array by Induced Sorting

### 2017-08-10 new-draft/draft.tex
2017/08/10

\title{Building and Checking Suffix and LCP Arrays Using Induced Sorting Method}

@@

Suffix and longest common prefix~(LCP) arrays can be built by the induced sorting~(IS) method on both internal and external memory models. We propose two methods that enable any IS builder to build and check suffix and LCP arrays simultaneously. The first method is for checking both suffix and LCP arrays, while the second method is for checking SA only. By combining the Karp-Rabin fingerprinting techinque into our methods, we design two algorithms that perform checking correctly with a negligible error probability. Theoretically, the algorithm designed by the first method has a sorting complexity, while the algorithm designed by the second method only takes linear time to run. We integrate the algorithm designed by the second method into the existing disk-based suffix sorting algorithm DSA-IS and implement their combination for performance evaluation. From our experiments, the checking overhead is considerably smaller than the building consumption.

@@

@@

A suffix array~(SA) can be built within linear time and space by the internal-memory algorithm SA-IS~\cite{Nong11}. According to the IS principle, the order of two suffixes is determined by comparing their heading characters and successors in sequence, where the order of their successors is assumed to be determined in advance. Recently, the IS method has been also applied to designing three disk-based suffix sorting algorithms eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}, where eSAIS can build both suffix and LCP arrays at the same time. These works have better time complexities than the other alternatives ~(e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}), but they all suffer from a bottleneck due to the large disk space for obtaining the heading characters of unsorted suffixes and the ranks of their sorted successors in a disk-friendly way. It is reported that the average peak disk use to construct an SA encoded by 40-bit integers for pSAscan is only $7.5n$, while that for eSAIS, DSA-IS and SAIS-PQ are $24n$, $18n$ and $15n$, respectively. The poor space performance of the IS algorithms is mainly because that their current programs fail to free the disk space for temporary data even when the data is no longer to use. A dramatic improvement can be achieved by storing temporary data in multiple files and deleting each file when the data residing on it is not needed any more. This technique has been used to implement a novel IS suffix sorting algorithm fSAIS~\cite{Karkkainen2017}, where the engineering of fSAIS takes no more than $8n$ disk space. This indicates a great potential for optimizing the performances of DSA-IS and SAIS-PQ.

A constructed SA should be checked to detect potential errors caused by implementation bugs and other malfunctions. The software packages for DC3 and eSAIS provide users a checker based on the idea presented in~\cite{Dementiev2008a}. When running on external-memory model, the cost taken by this checker is rather high, because it performs two passes of integer sorts for arranging $\mathcal{O}(n)$ fixed-size tuples. In this paper, we describe two methods that enable any IS builder to build and check suffix and LCP arrays simultaneously. The first method is for checking both suffix and LCP arrays, while the second method is for checking SA only. We employ the Karp-Rabin fingerprinting technique~\cite{Karp1987} to design two probabilistic algorithms, in terms of that their checking results are wrong with a negligible probability. For analysis, we first use new substring sorting and naming methods to improve the design of DSA-IS and then combine the second checking method into the adapted DSA-IS for evaluating the checking overhead. Our experimental results indicate that the time, space and I/O volume for checking SA is considerable smaller than that for building SA. This implies that the IS method can be used to design efficient solutions for SA verification as well.

@@

Partition $x$ into multiple S*-type substrings. Each substring only contains two S*-type characters and any two neighboring substrings overlap an S*-type character. Produce a reduced string $x_1$ by replacing each S*-type substring with its name, which represents the rank of the corresponding S*-type substring among all. In the following paragraphs, we use "rank" and "name" interchangeably to indicate the lexical order of the corresponding substring.

@@

All the suffixes in $sa$ are naturally grouped into multiple buckets. Each bucket occupies a contiguous interval in $sa$ and contains all the suffixes with a same heading character. Without loss of generality, we denote by ${\sf sa\_bkt}(c_0)$ the bucket for suffixes starting with $c_0$. It should be noticed that ${\sf sa\_bkt}(c_0)$ can be divided into two parts where the left part ${\sf sa\_bkt_L}(c_0)$ and the right part ${\sf sa\_bkt_S}(c_0)$ contain the L-type and S-type suffixes, respectively. We also define ${\sf lcp\_bkt}(c_0)$, ${\sf lcp\_bkt_L}(c_0)$ and ${\sf lcp\_bkt_S}(c_0)$ on $lcp$ for each character in $\Sigma$.

@@

Algorithm~\ref{alg:1} shows the framework of any IS suffix sorting algorithm. In lines~\ref{alg:1:ref:1}-\ref{alg:1:ref:2}, a reduction phase for sorting and naming S*-type substrings is called to produce the reduced string $x_1$. If there exist duplicate characters in $x_1$, the reduction phase is called with $x_1$ as input at the higher recursion level in line~\ref{alg:1:ref:3}; otherwise, all the S*-type suffixes in $x$ are already sorted and $sa_1$ is directly computed from $x_1$ in line~\ref{alg:1:ref:4}. Afterward, an induction phase for sorting suffixes is called to produce $sa$ for $x$ at the current recursion level in line~\ref{alg:1:ref:5}. The induction phase is recursively called with $sa$ as input at the lower recursion level (if any).

During the execution of a reduction/induction phase, all the substrings/suffixes are sorted by comparing their heading characters and the ranks of their successors according to the IS principle. This involves a great number of random accesses to $x$ and $sa$, which can be done very fast if both $x$ and $sa$ can wholly reside on RAM. However, if the size of input and output exceeds the capacity of internal memory, each access will take an individual I/O operation, leading to a performance degradation. The DSA-IS algorithm solves the problem by performing the following two steps during a reduction/induction phase:

@@

	\item Sort the substrings/suffixes by their heading characters and the ranks of their successors in an external-memory heap. Scan the sorted substrings/suffixes in the heap and induce their predecessors into the heap, where the heading characters of the induced substrings/suffixes have been retrieved from external memory in advance by sequential I/O operations. \label{dsais_sorting_method:2}
\end{enumerate}

As shown in Section~\ref{sec:experiments}, our program for DSA-IS requires less disk space than that for eSAIS, but the former runs slower than the latter due to the large I/O volume for sorting and naming S*-type substrings during the reduction phase. To improve the performance, we propose new substring sorting and naming methods in the following.

@@

\begin{enumerate}[S1']
	\item Sort the long in each block by S\ref{dsais_sorting_method:1}. During the process, copy the short to external memory in their sorted order.~\label{dsaism_sorting_method:1}

	\item Sort the long in $x$ by S\ref{dsais_sorting_method:2}. During the process, copy the leftmost $D$ characters of the long to external memory in their sorted order.~\label{dsaism_sorting_method:2}

	\item Merge the short and long by a multi-way sorter.~\label{dsaism_sorting_method:3}
\end{enumerate}

After S\ref{dsaism_sorting_method:1}'-S\ref{dsaism_sorting_method:2}', the short S*-type substrings in each block and the long S*-type substrings in $x$ are sorted and separately organized as a sequence in external memory. Assume $x$ is split into $k$ blocks, the multi-way sorter in S\ref{dsaism_sorting_method:3}' maintains an internal-memory heap to cache the current smallest of each sequence and continually retrieve the top item from the heap to determine the lexical order of substrings from different sequences. The heap contains at most $k + 1$ substrings at any point of time and it compares any two substrings in $\mathcal{O}(D)$ time by conducting a literal string comparison\footnote{If the leftmost $D$ characters of a long S*-type substring is equal to a short S*-type substring, then the short is lexicographically greater than the long.}. The above sorting method can achieve a good performance if the majority of S*-type substrings in $x$ are short, provided $D$ is small. This is commonly satisfied in real-world datasets.

Next, we describe a method for naming the S*-type substrings when they are being sorted. The key point here is to check equality of two substrings successively popped from the heap in S\ref{dsaism_sorting_method:3}' immediately. If either of the two substrings is short, then we literally compare them in $\mathcal{O}(D)$ time to check their equality. Otherwise, the two substrings are both long, we first determine their names when inducing them in S\ref{dsaism_sorting_method:2}' and compare these names in $\mathcal{O}(1)$ time to check their equality. This naming technique was originally proposed for SAIS-PQ for merging the sorting and naming processes into a whole. The experimental results in Section~\ref{sec:experiments} indicate that, by using the new substring sorting and naming methods, the adapted DSA-IS, called DSA-IS+, only takes half as much disk space as eSAIS.

@@

\begin{IEEEproof}The first condition indicates that $sa$ is a permutation of all the suffixes in $x$. The second condition indicates that the order of suffixes in $sa$ corresponds to that of their heading characters and successors. Because any two suffixes can be sorted by comparing their heading characters and successors, the above conditions are sufficient and necessary for SA verification.

@@

The disk-based implementation of this checker conducts two passes of integer sorts and each sort arranges the order of $\mathcal{O}(n)$ fixed-size tuples using external memory. As can be observed from Section~\ref{sec:experiments}, the peak disk use and the I/O volume for an SA encoded by 40-bit integers are around $26n$ and $53n$, respectively.

@@

Recall that, an IS suffix sorting algorithm recursively perform the reduction phase to compute the reduced string $x_1$ until $x_1$ contains no duplicate characters. Afterward, it produces $sa_1$ from $x_1$ and recursively perform the induction phase to compute $sa$ until reaching the top recursion level, where the induction phase consists of the following steps:

@@

A running example of S\ref{induction_phase:2}''-S\ref{induction_phase:4}'' is shown in Fig.~\ref{fig:example1}. Given $sa^*$ is already known, line 6 inserts each S*-type suffix into the right part of the corresponding bucket and arranges them according to their sorted order indicated by $sa^*$. For example, ${\sf suf}(8)$, ${\sf suf}(5)$, ${\sf suf}(2)$ are placed into the right part of ${\sf sa\_bkt}(i)$ and arranged in their sorted order indicated by $sa^*$. Next, we find the leftmost position of each bucket (marked by the symbol $\wedge$) and scan $sa$ rightward for inducing the order of L-type suffixes. For this, we first check $sa[0] = 14$ (marked by the symbol @) and induce the predecessor of ${\sf suf}(14)$ in lines 8-9. Because $x[13] = i$ is L-type, we put ${\sf suf}(13)$ into the current leftmost empty position in ${\sf sa\_bkt_L}(i)$. To step through $sa$ in this way, we get all the L-type suffixes sorted in line 17. Afterward, we find the rightmost position of each bucket and scan $sa$ leftward for inducing the order of S-type suffixes. When scanning $sa[14] = 4$ in lines 19-20, we see $x[3] = i$ is S-type and thus put ${\sf suf}(3)$ into the current rightmost empty position in ${\sf sa\_bkt_S}(i)$. Following the same idea, we get all the S-type suffixes sorted in line 28. Based on the discussion, we show in Lemma~\ref{lemma:2} a set of sufficient and necessary conditions for SA verification.

@@

It should be noticed that, we can check the correctness of $sa^*$ instead of $sa$ at the top recursion level to perform SA verification if the first condition of Lemma~\ref{lemma:2} is always true. In fact, the code snippet for the induction phase at the top recursion level only consists of tens of C++ code lines. This is considerable smaller than the whole program for a disk-based suffix sorting algorithm. Following the idea, we assume S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' are correctly implemented and propose two methods for checking computation errors caused by implementation bugs and other malfunctions. The algorithms designed by these methods can be seamlessly integrated into any IS suffix sorting algorithm for building and checking an SA at the same time.

@@

Method A is based on Lemma~\ref{lemma:3}, which is extended from Lemma~\ref{lemma:2}.

\begin{lemma} \label{lemma:3}
	Assume the induction phase is correctly implemented at the top recursion level, the output $sa[0, n)$ of an IS suffix sorting algorithm is the SA for the input string $x[0, n)$ if and only if the following conditions are satisfied: \\
	(1) $sa^*$ is correct. \\
	(2) $sa[i]$ is equal to the value calculated by S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' for $i \in [0, n)$. \\

@@

The first condition of Lemma~\ref{lemma:3} can be checked by a sparse SA checker like~\cite{wu2017}. The second condition intends for detecting malfunctions other than implementation bugs (e.g., I/O errors). This is checked by determining whether the induced and scanned values for each suffix are equal\footnote{Notice that each suffix induced into $sa$ will be latter scanned for inducing the order of its predecessor during S\ref{induction_phase:3}'' and S\ref{induction_phase:4}''. }. The problem to be solved here is that when a suffix is induced into a bucket, its corresponding value in $sa$ may not be scanned at once. Our solution is to integrate the checking process into the building process, by ensuring the sequence of values placed at a bucket is identical to that scanned later. For the purpose, we use a fingerprinting function to increasingly compute the fingerprints of both sequences and check their equality in constant time at the end of the induction phase. If the two fingerprints for each bucket are equal, then the second condition of Lemma~\ref{lemma:3} will be seen with a high probability. As a result, $sa$ can be built and probabilistically checked at the same time. It should be noticed that this method can be also applied to checking the LCP array when it is being built from $lcp^*$ following the IS principle~\cite{Fischer11}. As a result, we design Algorithm~\ref{alg:2} to probabilistically check $sa$ and $lcp$ at the top recursion level. In the algorithm, ${\sf sa_{I1}}(c)$ and ${\sf sa_{I2}}(c)$ are two sequences respectively induced into ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$, while ${\sf sa_{S1}}(c)$ and ${\sf sa_{S2}}(c)$ are two sequences respectively scanned from ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$.

@@

\begin{lemma} \label{lemma:4}
    Assume the induction phase is correctly implemented at the top recursion level, the output $sa[0, n)$ of an IS suffix sorting algorithm is the SA for the input string $x[0, n)$ if and only if the following conditions are satisfied:

@@

    We only prove the sufficiency as the necessity is clear.

    Case 1: Suppose all the S*-type suffixes are correctly sorted in $sa^*$, then $sa$ is right according to Lemma~\ref{lemma:3}.

@@

The first condition of Lemma~\ref{lemma:4} is naturally satisfied when computing $sa^*$ from $sa_1$ in S\ref{induction_phase:1}''. The second condition is probabilistically checked by computing and comparing the fingerprints of $sa^*$ and $\overline{sa^*}$. The fingerprint of $sa^*$ is computed when placing the elements of $sa^*$ into $sa$ in S\ref{induction_phase:2}'', while the fingerprint of $\overline{sa^*}$ is computed when inducing the S*-type suffixes into $sa$.

@@

@@

The proposed two methods check equality of two arrays (or sub-arrays) by comparing their fingerprints, which can be calculated using a hash function. Notice that two equal arrays must share an identical hash value, but the inverse is not always true. To lower the error probability of a false match, we prefer using the Karp-Rabin fingerprinting function to compute the required values following Formulas~\ref{formula:1}-\ref{formula:2}, where $L$ is a prime and $\delta$ is an integer randomly chosen from $[1, L)$. By setting $L$ to a large value, the error probability can be reduced to a negligible level. In Fig.~\ref{fig:example2}, we depict an example for computing the integer array $A$, where $A$ is identical to $sa^*$ and $\overline{sa^*}$ in Fig.~\ref{fig:example1}. Given $L = 23$ and $\delta = 5$, we first compute the fingerprint of $A[0, 0]$ in line 4. Because $\sf fp(A[0, -1]) = 0$, $\sf fp(A[0, 0]) = (0 \cdot 5 + 14) \, mod \, 23 = 14$. By iteratively computing the fingerprints of the prefixes of $A$, we finally obtain the fingerprint of $A$ in line 7.

@@

Assuming the induction phase at the top recursion level is correctly implemented, the bottleneck of Algorithms~\ref{alg:2} and~\ref{alg:3} occur when checking $sa^*$. For Algorithm~\ref{alg:2}, we apply the method proposed in~\cite{wu2017} to ensure the correctness of $sa^*$ within the sorting complexity. Specifically, given two suffixes, i.e. ${\sf suf}(i)$ and ${\sf suf}(j)$, and their LCP-value $\ell$, we compute and compare the fingerprints of ${\sf sub}(i, i + \ell - 1)$ and ${\sf sub}(j, j + \ell - 1)$. If the fingerprints are equal and the order of $x[i + \ell]$ and $x[j + \ell]$ corresponds to that of ${\sf suf}(i)$ and ${\sf suf}(j)$, then the suffixes are correctly sorted and their LCP-value is right with a high probability. Because at most one out of every two suffixes is S*-type (commonly one-third in real-world data sets) and the checking process is only executed during the induction phase at the top recursion level, the checking overhead is much less than the building consumption. This conclusion is also true for Algorithm~\ref{alg:3}, where the checking overhead is mainly taken by computing fingerprints in need.

>>>>

\title{Building and Checking a Suffix Array by Induced Sorting}

@@

A suffix array can be built by the induced sorting~(IS) method on both internal and external memory models. We propose two methods for checking a suffix array when it is being built by any IS suffix sorter, i.e. build and check a suffix array by the IS method simultaneously. In particular, the first method can be extended to check both suffix and LCP arrays. By combining the Karp-Rabin fingerprinting technique with our methods, we design two algorithms that perform checking correctly with a negligible error probability. The algorithm designed by the first method has an integer sorting complexity, while the algorithm designed by the second method runs in linear time. We integrate the algorithm designed by the second method into the existing disk-based suffix sorting algorithm DSA-IS and implement it for performance analysis. From our experiments, the checking overhead is considerably smaller than the building consumption.

%A suffix array can be built by the induced sorting~(IS) method on both internal and external memory models. We propose two methods to enable the IS to check a suffix array when it is built, i.e. build and check a suffix array simultaneously. In particular, the first method can be exteneded to check both suffix and LCP arrays. By combining the Karp-Rabin fingerprinting techinque with our methods, we design two algorithms that can check correctly with a negligible error probability. The algorithm designed by the first method has an integer sorting complexity, while the algorithm designed by the second method runs in linear time. We integrate the algorithm designed by the second method into the existing disk-based suffix sorting algorithm DSA-IS and implement it for performance evaluation. From our experiments, the checking overhead is considerably smaller than the building consumption.

@@

{\color{red}

In the accepted paper, the LCP-value $\ell$ of ${\sf suf}(i)$ and ${\sf suf}(j)$ is assumed to be right if ${\sf sub}(i, i + \ell - 1) = {\sf sub}(j, j + \ell - 1)$ and $x[i + \ell] \ne x[j + \ell]$. Given that $\ell$ is correct, the lexical order of ${\sf suf}(i)$ and ${\sf suf}(j)$ is checked by comparing $x[i + \ell]$ and $x[j + \ell]$.

In this paper, both the lexical order and the LCP-value of two neighboring suffixes in the given suffix array is checked at the top recursion level following the IS principle.

}

@@

A suffix array~(SA) can be built within linear time and space by the internal-memory algorithm SA-IS~\cite{Nong11} using the induced sorting method. According to the IS principle, the order of two suffixes is determined by comparing their heading characters and successors in sequence, where the order of their successors is determined recursively. Recently, the IS method has been also applied to designing three disk-based suffix sorting algorithms eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}, where eSAIS can build both suffix and LCP arrays at the same time. Compared to the others designed by different methods (e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}), these algorithms runs fast with linear I/O complexity. However, these algorithms currently suffer from a bottleneck due to the large disk space for obtaining the heading characters of unsorted suffixes and the ranks of their sorted successors in a disk-friendly way. It is reported that the average disk use for constructing an SA encoded by 40-bit integers for pSAscan is only $7.5n$ bytes, while that for eSAIS, DSA-IS and SAIS-PQ are $24n$, $18n$ and $15n$ bytes, respectively, where $n$ is the size of input string. The poor space performance observed from these IS algorithms is mainly because the current programs for these algorithms fail to immediately free the disk space for temporary data even when the data is no longer needed. A dramatic improvement can be achieved by storing temporary data in multiple files and deleting a file when the data in the file is not needed any more. This trick has been used by fSAIS~\cite{Karkkainen2017} to engineer the external-memory IS method for better performance, where the peak disk use is $8n$ bytes for constructing an SA of 40-bit integers in the given experiments.

While the research on developing various SA builders using the IS method is involving, the developed software are becoming complex and appear as open-source without guarantee. A constructed SA should be checked to detect potential errors caused by implementation bugs and other malfunctions. The software packages for DC3 and eSAIS provide a checker based on the idea presented in~\cite{Dementiev2008a}. When running on external-memory model, the cost taken  by this checker is rather high as it performs two passes of integer sorts for arranging $\mathcal{O}(n)$ fixed-size tuples. In this paper, we propose two methods to enable the IS method to build and check an SA simultaneously. In particular, the first method can be extended to check both suffix and LCP arrays. We employ the Karp-Rabin fingerprinting technique~\cite{Karp1987} to design two probabilistic algorithms, in terms of that their checking results are wrong with a negligible probability. For analysis, we first use new substring sorting and naming methods to improve the design of DSA-IS and then combine the second checking method into the adapted DSA-IS for evaluating the checking overhead. Our experimental results indicate that the time, space and I/O volume for checking SA is considerably smaller than that for building SA.

@@

Partition $x$ into multiple S*-type substrings, each substring contains two S*-type characters and any two neighboring substrings overlap an S*-type character. We produce a reduced string $x_1$ by replacing each substring with its name, where the name indicates the rank of the	 substring among all. In this paper, we use "rank" and "name" interchangeably to represent the lexical order of substrings.

@@

All the suffixes in $sa$ are naturally grouped into multiple buckets. Each bucket occupies a contiguous interval in $sa$ and contains all the suffixes with a same heading character. Without loss of generality, we denote by ${\sf sa\_bkt}(c_0)$ the bucket for suffixes starting with $c_0$, where ${\sf sa\_bkt}(c_0)$ can be naturally divided into two parts ${\sf sa\_bkt_L}(c_0)$ and ${\sf sa\_bkt_S}(c_0)$ containing L-type and S-type suffixes, respectively. We also define ${\sf lcp\_bkt}(c_0)$, ${\sf lcp\_bkt_L}(c_0)$ and ${\sf lcp\_bkt_S}(c_0)$ on $lcp$ for $c_0 \in \Sigma$.

@@

Algorithm~\ref{alg:1} shows the framework of an IS suffix sorting algorithm. In lines~\ref{alg:1:ref:1}-\ref{alg:1:ref:2}, a reduction phase for sorting and naming S*-type substrings is executed to produce the reduced string $x_1$. If there exist duplicate characters in $x_1$, then the reduction phase is recursively called with $x_1$ as input in line~\ref{alg:1:ref:3} and executed at the higher recursion level; otherwise, the S*-type suffixes in $x$ are already sorted and $sa_1$ is directly computed from $x_1$ in line~\ref{alg:1:ref:4}. Afterward, an induction phase for sorting suffixes is executed to produce $sa$ at the current recursion level in line~\ref{alg:1:ref:5}. The induction phase is called recursively with $sa$ as input and executed at the lower recursion level until the final $sa$ is generated.

During the execution of a reduction/induction phase, all the substrings/suffixes are sorted by comparing their heading characters and the ranks of their successors according to the IS principle. This involves a great number of random accesses to $x$ and $sa$, which can be done very fast if both $x$ and $sa$ can wholly reside on RAM. However, if the size of input and output exceeds the capacity of internal memory, each access will take an individual I/O operation to disk, leading to a performance degradation. The DSA-IS algorithm solves this problem by performing the following two steps during a reduction/induction phase:

@@

	\item Sort the substrings/suffixes by their heading characters and the ranks of their successors in an external-memory heap. \label{dsais_sorting_method:2}
\end{enumerate}

Because all the substrings/suffixes in the heap are sorted already, S\ref{dsais_sorting_method:2} determines the rank of a substring/suffix when popping it from the heap. When inducing the predecessor of current top substring/suffix, the heading character in need is retrieved from external memory by a sequential I/O operation. As shown in Section~\ref{sec:experiments}, our program for DSA-IS requires less disk space than that for eSAIS, but the former runs slower than the latter due to the large I/O volume for sorting and naming S*-type substrings during the reduction phase. To improve the performance, we propose new substring sorting and naming methods in the following.

@@

\begin{enumerate}[S1']
	\item Sort the long in each block, copy the heading characters in need to external memory by S\ref{dsais_sorting_method:1}. Meanwhile, sort the short in each block, copy the short to external memory in their sorted order.~\label{dsaism_sorting_method:1}

	\item Sort the long in $x$ by S\ref{dsais_sorting_method:2}, copy the leftmost $D$ characters of the long to external memory in their sorted order.~\label{dsaism_sorting_method:2}

	\item Merge the short and the long by a multi-way sorter.~\label{dsaism_sorting_method:3}
\end{enumerate}

The sorted short S*-type substrings in each block are organized as a sequence in external memory after S\ref{dsaism_sorting_method:1}', while the sorted long S*-type substrings in $x$ are also organized as a sequence in external memory after S\ref{dsaism_sorting_method:2}'. Assume $x$ is split into $k$ blocks, the multi-way sorter in S\ref{dsaism_sorting_method:3}' maintains an internal-memory heap to determine the lexical order of substrings belonging to different sequences. The heap contains at most $k + 1$ substrings at any point of time. It compares any two substring in $\mathcal{O}(D)$ time by conducting a literal string comparison\footnote{If the leftmost $D$ characters of a long S*-type substring is equal to a short S*-type substring, then the short is lexicographically greater than the long.}.

The above sorting method has a good performance if $D$ is small and the majority of S*-type substrings in $x$ are short. This is commonly satisfied in real-world datasets. We next describe a method for naming the S*-type substrings when they are being sorted according to S\ref{dsaism_sorting_method:1}'-S\ref{dsaism_sorting_method:3}'. The key point is to check equality of two suffixes successively popped from the heap. For this, we literally compare the two substrings in $\mathcal{O}(D)$ time if either of them is short; otherwise, we compare their names in $\mathcal{O}(1)$ time instead. Recall that the rank of each long S*-type substring is determined when inducing its predecessor, this value is reused to represent its name herein\footnote{A similar technique has been used to merge the substring sorting and naming process in SAIS-PQ.}. The experimental results in Section~\ref{sec:experiments} demonstrate that, by using the new substring sorting and naming methods, the adapted DSA-IS, called DSA-IS+, only takes half as much disk space as eSAIS.

@@

\begin{IEEEproof} If condition (1) is true, then $sa$ is a permutation of all the suffixes in $x$. If condition (2) is true, then the lexical order of any two neighboring suffixes in $sa$ is determined by comparing their heading characters and the ranks of their successors. As a result, all the suffixes in $x$ are sorted according to the lexical order of their heading characters and successors.

@@

The disk-based implementation of this checker conducts two passes of integer sorts and each sort arranges the order of $\mathcal{O}(n)$ fixed-size tuples in external memory. As can be observed from Section~\ref{sec:experiments}, the peak disk use and the I/O volume for an SA encoded by 40-bit integers are around $26n$ and $53n$, respectively.

@@

Recall that, an IS suffix sorting algorithm recursively executes the reduction phase to compute the reduced string $x_1$ until $x_1$ contains no duplicate characters. Afterward, it produces $sa_1$ from $x_1$ and recursively executes the induction phase to compute $sa$ until reaching the top recursion level, where the induction phase consists of the following steps:

@@

A running example of S\ref{induction_phase:2}''-S\ref{induction_phase:4}'' is shown in Fig.~\ref{fig:example1}. Given $sa^*$ is already known, line 6 inserts each S*-type suffix into the corresponding bucket. For example, ${\sf suf}(2)$, ${\sf suf}(5)$, ${\sf suf}(8)$ are sequentially placed into the current rightmost empty position of ${\sf sa\_bkt}(i)$, where the insertion order corresponds to their sorted order indicated by $sa^*$. The next step is to find the leftmost position of each bucket (marked by the symbol $\wedge$) and scan $sa$ rightward for inducing the order of L-type suffixes. For this, we first check $sa[0] = 14$ (marked by the symbol @) and induce the predecessor of ${\sf suf}(14)$ in lines 8-9. Because $x[13] = i$ is L-type, we put ${\sf suf}(13)$ into the current leftmost empty position in ${\sf sa\_bkt_L}(i)$. To step through $sa$ in this way, we get all the L-type suffixes sorted in line 17. Afterward, we find the rightmost position of each bucket and scan $sa$ leftward for inducing the order of S-type suffixes. When scanning $sa[14] = 4$ in lines 19-20, we see $x[3] = i$ is S-type and thus put ${\sf suf}(3)$ into the current rightmost empty position in ${\sf sa\_bkt_S}(i)$. Following the same idea, we get all the S-type suffixes sorted in line 28. Following the discussion, we show in Lemma~\ref{lemma:2} a set of sufficient and necessary conditions for checking SA.

@@

Lemma~\ref{lemma:2} shows that we can check the correctness of $sa^*$ instead of $sa$ at the top recursion level to perform SA verification if the first condition of Lemma~\ref{lemma:2} is always true. In fact, the code snippet for the induction phase at the top recursion level only consists of tens of C++ code lines. This is considerably smaller than the whole program for a disk-based suffix sorting algorithm. Following the idea, we assume S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' are correctly implemented and propose two methods for checking computation errors caused by implementation bugs and other malfunctions. The algorithms designed by these methods can be seamlessly integrated into any IS suffix sorting algorithm for building and checking an SA at the same time.

@@

Method A is based on Lemma~\ref{lemma:3}, which is an extension of Lemma~\ref{lemma:2}.

\begin{lemma} \label{lemma:3}
	Assume the induction phase at the top recursion level is correctly implemented, the output $sa[0, n)$ of an IS suffix sorter is the SA for the input string $x[0,  n)$ if the following conditions are satisfied: \\
	(1) $sa^*$ is correct. \\
	(2) $sa[i]$ is equal to the value calculated by S\ref{induction_phase:1}''-S\ref{induction_phase:4}'' for $i \in [0, n)$. \\

@@

The first condition of Lemma~\ref{lemma:3} can be checked by a sparse SA checker like~\cite{wu2017}. The second condition intends for detecting malfunctions other than implementation bugs (e.g., I/O errors). This is checked by determining whether the induced and scanned values for each suffix are equal\footnote{Notice that each suffix induced into $sa$ will be latter scanned for inducing the order of its predecessor during S\ref{induction_phase:3}'' and S\ref{induction_phase:4}''. }. The problem to be solved here is that when a suffix is induced into a bucket, its corresponding value in $sa$ may not be scanned at once. Our solution is to ensure the sequence of values placed at a bucket is identical to that scanned latter. For the purpose, we use a fingerprinting function to increasingly compute the fingerprints of both sequences and check their equality in constant time at the end of the induction phase. If the two fingerprints for each bucket are equal, then the second condition of Lemma~\ref{lemma:3} will be seen with a high probability. As a result, $sa$ can be built and probabilistically checked at the same time.

It was reported in~\cite{Fischer11, Bingmann12} that the IS method can be used to construct LCP array as well. For any two neighboring suffixes in $sa$, their LCP-value is zero if they start with different characters; otherwise, the LCP-value of them is one greater than that of their successors. This implies the possibility for extending Method A to check both suffix and LCP arrays. We design Algorithm~\ref{alg:2} to check $sa$ and $lcp$ when they are induced from $sa^*$ and $lcp^*$ at the top recursion level of an IS builder. In this algorithm, ${\sf sa_{I1}}(c)$ and ${\sf sa_{I2}}(c)$ are two sequences respectively induced into ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$, while ${\sf sa_{S1}}(c)$ and ${\sf sa_{S2}}(c)$ are two sequences respectively scanned from ${\sf sa\_bkt_L}(c)$ and ${\sf sa\_bkt_S}(c)$. These values are used to check the second condition of Lemma~\ref{lemma:3} later. We also compute ${\sf lcp_{I1}}(c)/{\sf lcp_{I2}}(c)$ and ${\sf lcp_{S1}}(c)/{\sf lcp_{S2}}(c)$ and check their equality in the meantime.

@@

\begin{lemma} \label{lemma:4}
    Assume the induction phase at the top recursion level is correctly implemented, the output $sa[0, n)$ of an IS suffix sorter is the SA for the input string $x[0, n)$ if the following conditions are satisfied:

@@

    We prove the sufficiency according to the following two cases.

    Case 1: Suppose all the S*-type suffixes are correctly sorted in $sa^*$, then $sa$ is right based on Lemma~\ref{lemma:3}.

@@

The first condition of Lemma~\ref{lemma:4} is naturally satisfied when computing $sa^*$ from $sa_1$ in S\ref{induction_phase:1}''. The second condition is probabilistically checked by computing and comparing the fingerprints of $sa^*$ and $\overline{sa^*}$. The fingerprint of $sa^*$ is computed when placing the elements of $sa^*$ into $sa$ in S\ref{induction_phase:2}'', while the fingerprint of $\overline{sa^*}$ is computed when inducing the S*-type suffixes into $sa$ in S\ref{induction_phase:4}''. We design Algorithm~\ref{alg:3} to check if the final SA observes Lemma~\ref{lemma:4}.

@@

The proposed two methods check equality of two arrays (or sub-arrays) by comparing their fingerprints, which can be calculated using a hash function. Notice that two equal arrays must be mapped to an identical hash value, but the inverse is not always true. To lower the error probability of a false match, we prefer using the Karp-Rabin fingerprinting function to compute these hash values, called fingerprints,	 following Formulas~\ref{formula:1}-\ref{formula:2}, where $L$ is a prime and $\delta$ is an integer randomly chosen from $[1, L)$. By setting $L$ to a large value, the error probability can be reduced to a negligible level. In Fig.~\ref{fig:example2}, we depict an example for computing the integer array $A$ identical to $sa^*$ and $\overline{sa^*}$ in Fig.~\ref{fig:example1}. Given $L = 23$ and $\delta = 5$, we first compute the fingerprint of $A[0, 0]$ in line 4. Because $\sf fp(A[0, -1]) = 0$, $\sf fp(A[0, 0]) = (0 \cdot 5 + 14) \, mod \, 23 = 14$. By iteratively computing the fingerprints of the prefixes of $A$, we finally obtain the fingerprint of $A$ in line 7.

@@

@@

Assume the induction phase at the top recursion level is correctly implemented, Algorithm~\ref{alg:2} employs the method proposed in~\cite{wu2017} to ensure the correctness of $sa^*$ and $lcp^*$. To be specific, for two neighboring suffixes ${\sf suf}(i)$ and ${\sf suf}(j)$ in $sa^*$, their LCP-value $\ell$ in $lcp^*$ is right if ${\sf sub}(i, i + \ell - 1) = {\sf sub}(j, j + \ell - 1)$ and $x[i + \ell] \ne x[j + \ell]$, where the equality of ${\sf sub}(i, i + \ell - 1)$ and ${\sf sub}(j, j + \ell - 1)$ is checked by comparing the fingerprints of these substrings. Given that $\ell$ is correct, the lexical order of ${\sf suf}(i)$ and ${\sf suf}(j)$ is checked by comparing $x[i + \ell]$ with $x[j + \ell]$. This method computes the fingerprints of all the target substrings within a sorting complexity proportional to $n_1$. Because at most one out of every two suffixes is S*-type, the checking overhead is much less than building consumption. By contrast, Algorithm~\ref{alg:3} takes overall $\mathcal{O}(n)$ to compute all the required fingerprints.
